\textbf{\sc Towards Deliberative Control of Autonomous Robots}

### 2012-02-06 trunk/concepts.tex
updates

Our system, \texttt{T-REX} (or the Teleo-Reactive EXecutive) is an
adaptive, artificial intelligence based controller and provides
general framework for building reasoning systems for real-world
autonomous vehicles. At MBARI \texttt{T-REX} is used for AUV control;
another instantiation of the system is being used for control of a
terrestrial personal robot \cite{pr2, Meeussen:2010dn}. The development of
\texttt{T-REX} has been targeted at surveying a number of
oceanographic features which are dynamic and unpredictable
spatio-temporally. Typically this requires our AUVs to balance the
goals of coverage spatially while opportunistically following features
of scientific interest and to do so while being operationally aware of
its own limitations in terms of resources (typically the battery state
of charge) and overall proximity to other observational assets for
obtaining scientific ground-truth. We are therefore interested in
representational frameworks that allow robots to pursue long-term
objectives while choosing short-term gain and can gracefully deal with
exogenous or endogenous change.

To enable this responsiveness to external observations, the agent has
to be able to synthesize plans insitu and to re-plan. We use a
temporal constraint-based planner EUROPA$_2$ with a demonstrated
legacy of having flown on NASA space missions
\cite{Jonsson00,bresina05}. Our autonomy architecture brings three key
innovations for AUV adaptation: the use of flexible plan
representations, compositional control with the use of partitioned
networks and off-line learning to inform insitu environmental state
estimation. In this section we detail some of the key concepts
relevant to flexible plan representation. Section \ref{sec:arch} has
details the \texttt{T-REX} partitioned architecture and estimation
driven vehicle adaptation.

% \texttt{T-REX} uses EUROPA$_2$ a constraint-based temporal planner
% \cite{Jonsson00,Frank2003} developed for NASA missions, most recently
% for the Mars Exploration Rovers \cite{bresina05}.

EUROPA$_2$ uses a domain model written in a declarative language
(NDDL: New Domain Description Language), together with initial
conditions and goals also specified in NDDL, to construct a set of
temporal relations that must be true at the start time. These models
include assertions about the physics of the vehicle, i.e how it
responds to external stimulus and internally driven goals. By
propagating these relations forward using Simple Temporal Networks
\cite{dechter91} and applying goal constraints, EUROPA$_2$ can select
a set of conditions that should be true in the future, where some of
these conditions will correspond to actions the agent must take. The
planner can backtrack and try another path during search if a goal
cannot be reached while being capable of discarding unachievable
goals.

Traditionally robot execution has relied on dispatching commands at
precise times. Such linear sequences of precisely timed commands give
no ability to adjust execution on the basis of sensory
information. Although some commands can issue tests on sensor
readings, these tests have the objective of verifying whether expected
execution conditions are occuring. If not, the state of the system is
declared off-nominal and execution of the sequence of commands is
interrupted. More recently, executives have been proposed and
implemented that significantly broaden the way robots can be commanded
\cite{mus98,Alami:1998p820}. For example, the Remote Agent executive
interpreted a \textit{temporally flexible plan} which represents each
start time as a variable and contains an explicit network of bounded
delay constraints between such variables.

@@

EUROPA$_2$ uses a \emph{state variable} representation to describe the
evolution of state over time. The instantiated history of such state
variable evolution over a temporal horizon we call \emph{timelines}
and which represent a single thread in the execution of a concurrent
system. At any given time each thread can execute a single procedure.
Thus each timeline consists of a sequence of procedures which
encapsulate and describe state evolution; we call these instantiated
atomic entities \emph{tokens}.  A token therefore describes a
procedure invocation, the state variables on which it can occur, the
parameter values of the procedure, and the time values defining the
interval. We allow encapsulation of uncertainty within these tokens
with a range of start and end times and parameters, all of which are
encoded as variables. A constraint solver in turn manipulates these
variables defined in a EUROPA$_2$ domain model. For example, consider
a scientific need to take a water samples 100 meters from a hotspot
while an AUV is performing a Yo-Yo in the water-column. Two samples
are needed if the feature's signal is above a threshold; one
otherwise. The token that is capturing the sensory threshold has a
parametric constraints to the token which fires the requisite water
sampler. In addition, the start time of the water sampling procedure
is highly dependant on the variability of sub-sea currents and actual
speed of the vehicle. Therefore a number of values are possible for
the start times and duration of the sampling all of which are valid
combinations for desired outcomes.

@@

@@

Each of the token variables, including the parameter variables, has a
domain of values assigned to it.  The variables may also participate
in constraints that specify which value combinations are valid.  To
allow the specification of multiple values, e.g, to express a range of
possible start times, variables are used to specify parameter, start
and end time values for a token.  As a result, a token $T$ is a tuple
$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
$s\leq e$) are variables encoded as flexible token start and end time
points. Consequently, partial plans represent a \textit{range of
  possible solutions} rather than an explicit definition of a single
trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
of a part of such plans with tokens representing constraints on
flexible temporal intervals. Systematic computation is reflected by
the use of the \texttt{Allen Algebra} \cite{allen84} temporal
relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
interval arithmetic in the EUROPA$_2$ domain model that allows
encoding of temporal constraint relationships between tokens on the
same or other concurrent timelines. Details of propagation algorithms
or of interval logic are beyond the scope of this paper.

@@

\begin{minipage}[c]{\textwidth}

% \framebox[\textwidth][t]{
Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
$\Rightarrow$ \\
met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
\ldots{} \\
ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
\end{minipage}

>>>>

\texttt{T-REX} is an adaptive, artificial intelligence based
controller and provides general framework for building reasoning
systems for real-world autonomous vehicles. At MBARI \texttt{T-REX} is
used for AUV control; another instantiation of the system is being
used for control of a terrestrial personal robot \cite{pr2,
  Meeussen:2010dn}. The development of \texttt{T-REX} has been
targeted at surveying a number of oceanographic features which are
dynamic and unpredictable spatio-temporally. Typically this requires
our AUVs to balance the goals of coverage spatially while
opportunistically following features of scientific interest and to do
so while being operationally aware of its own limitations in terms of
resources (typically the battery state of charge) and overall
proximity to other observational assets for obtaining scientific
ground-truth. We are therefore interested in representational
frameworks that allow robots to pursue long-term objectives while
choosing short-term gain and can gracefully deal with exogenous or
endogenous change.

To enable this responsiveness to external observations, the agent has
to be able to synthesize plans insitu and to re-plan. We use a
temporal constraint-based planner \eu with a demonstrated legacy of
having flown on NASA space missions \cite{jonsson00,bresina05,
  barreiro09}. Our autonomy architecture brings three key innovations
for AUV adaptation: the use of flexible plan representations,
compositional control with the use of partitioned networks and
off-line learning to inform insitu environmental state estimation. In
this section we detail some of the key concepts relevant to flexible
plan representation. Section \ref{sec:arch} has details the
\rx partitioned architecture and estimation driven vehicle
adaptation.

\subsection{The basics}
\label{sec:basics}

\eu uses a \emph{domain model} written in a declarative language
(NDDL: New Domain Description Language), together with initial
conditions and goals also specified in NDDL, to construct a set of
temporal relations that must be true at the start time. These models
include assertions about the physics of the vehicle, i.e how it
responds to external stimulus and internally driven goals. By
propagating these relations forward using Simple Temporal Networks
\cite{dechter91} and applying goal constraints, \eu can select a set
of conditions that should be true in the future, where some of these
conditions will correspond to actions the agent must take. The planner
can backtrack and try another path during search if a goal cannot be
reached while being capable of discarding unachievable goals.

Traditionally robot execution has relied on dispatching commands at
precise times. Such linear sequences of precisely timed commands give
no ability to adjust execution on the basis of sensory
information. Although some commands can issue tests on sensor
readings, these tests have the objective of verifying whether expected
execution conditions are occuring. If not, the state of the system is
declared off-nominal and execution of the sequence of commands is
interrupted. More recently, executives have been proposed and
implemented that significantly broaden the way robots can be commanded
\cite{mus98,alami:1998p820}. For example, the Remote Agent executive
interpreted a \textit{temporally flexible plan} which represents each
start time as a variable and contains an explicit network of bounded
delay constraints between such variables.

@@

@@

\eu uses a \emph{state variable} representation to describe the
evolution of state over time. The instantiated history of such state
variable evolution over a temporal horizon we call \emph{timelines}
and which represent a single thread in the execution of a concurrent
system. At any given time each thread can execute a single procedure.
Thus each timeline consists of a sequence of procedures which
encapsulate and describe state evolution; we call these instantiated
atomic entities \emph{tokens}.  A token therefore describes a
procedure invocation, the state variables on which it can occur, the
parameter values of the procedure, and the time values defining the
interval. We allow encapsulation of uncertainty within these tokens
with a range of start and end times and parameters, all of which are
encoded as variables. A constraint solver in turn manipulates these
variables defined in a \eu domain model. For example, consider
a scientific need to take a water samples 100 meters from a hotspot
while an AUV is performing a Yo-Yo in the water-column. Two samples
are needed if the feature's signal is above a threshold; one
otherwise. The token that is capturing the sensory threshold has a
parametric constraints to the token which fires the requisite water
sampler. In addition, the start time of the water sampling procedure
is highly dependant on the variability of sub-sea currents and actual
speed of the vehicle. Therefore a number of values are possible for
the start times and duration of the sampling all of which are valid
combinations for desired outcomes.

@@

\subsubsection{\eu Plan Representation}
\label{sec:europa:pr}

This section describes the fundamental entities within \eu which are
typical to other CAIP-based formalisms.

\begin{description}

\item[\textbf{Variables}] Values that need to be represented to
  describe the problem domain and over which we may want to specify
  constraints. In the Shopping Agent example, the times at which the
  agent needs to leave or be back, or executes a purchase, are all
  instances where variables representing time would be used.

\item[\textbf{Objects}] The items we wish to describe and refer to in
  a domain are considered Objects. As in the case with object-oriented
  analysis and design, one can seek out the nouns in any domain
  description to find likely objects. In the Shopping Agent example,
  we night consider the Agent itself, Products and Locations all to be
  objects. Objects have state and behavior. For example, a Shopping
  Agent can have:

  \begin{itemize}

  \item State: it's location, a bag that contains the products it has
    already purchased (the bag can in turn be another object), etc.

  \item Behavior: go to a location to look for a product, perform a
    purchase, return home, etc.

  \end{itemize}

  Similarly objects that have similar state and behavior can be
  described generically in terms of object types or classes. \eu
  allows the definition of object classes in the same way that it is
  done in popular object-oriented programming languages. However, to
  describe state and behavior for the purposes of planning we need to
  build on the formalism of first order logic as explained below.

\item[\textbf{Tokens}] In first order logic, a predicate defines a
  relation between objects and properties \comment{citation?}.  In
  \eu, we define such relations between variables whose domains are
  sets of objects and sets of properties to describe state and
  behavior. For example, we might use a predicate \texttt{At($a$,$l$)}
  to indicate that agent $a$ is at location $l$, or a predicate
  \texttt{Buy($a$,$p$}) to indicate that agent $a$ is taking acting to
  buy product $p$. Note that $a$ is a variable which may have a number
  of possible values in a problem with multiple agents. Similarly, $l$
  and $p$ are variables whose values are the set of possible positions
  and products respectively. \eu can be used to create partial plans,
  where the domains of variables $a$ and $l$ can have more than one
  possible value in them, or grounded plans, where single values will
  be specified for each variable as we saw in the case of the
  N-queen's problem.

  In general, to come up with an executable plan it is not sufficient
  to state predicates that describe required state or behavior without
  also specifying some temporal extent over which each of those
  predicates hold. Predicates that are always true can be thought to
  hold from the beginning to the end of time. However, in practice,
  the temporal extent of interest must be defined with timepoints to
  represent its start and end. So we might want to write
  \texttt{At($a$,$s$,$e$,$l$)} to indicate that the agent $a$ is at
  location $p$ from time $s$ to time $e$. In fact, this pattern of
  using such predicates to describe both state and behavior of objects
  is so prevalent in \eu that we have introduced a special construct
  called a Token which has the built in variables to indicate the
  object to which the statement principally applies and the timepoints
  over which it holds. A Token is an instance of a predicate that
  represents and object's state or behavior and is defined over a
  temporal extent. In \eu, all predicate instances are Tokens. Also,
  in the same way that objects are described generically by classes,
  in \eu Tokens can be described generically by Token Types. Every
  token has five built-in variables:

  \begin{enumerate}

  \item \textit {start}: The beginning of the temporal extent over
    which the predicate is defined.

  \item \textit {end}: The end of the temporal extent over which the
    predicate is defined.

  \item \textit {duration}: The constraint \textit{start} $+$
    \textit{duration} = \textit{end} is enforced automatically.

  \item \textit{object}: The set of objects to which a token might
    apply. In a grounded plan each Token applies to a specific Object,
    reflecting the intuition that we are using Tokens to describe some
    aspect of an Object (i.e. its state or behavior) in time. However,
    in a partial plan, the commitment to a specific object may not yet
    have been made.

  \item \textit{state}: Tokens can be \texttt{ACTIVE, INACTIVE,
      MERGED}, or \texttt{REJECTED}. The state variable captures the
    token's current state and its reachable states through further
    restriction.  This is \eus mechanism to support the CAIP approach.

\end{enumerate}

\comment {resume cleanup here}

% \textit{Objects (continued)}

% As we have noted, Tokens describe some aspect of an Object in
% time. Objects thus may have many Tokens in a plan in order to describe
% their state and behavior throughout all points in time of the
% plan. Within this general framework, we note a few particulars:

% Static Facts: Classes without predicates lead to Objects without
% Tokens. This arises where an objects state or behavior is independent
% of time. For example, a domain may have a set of locations and/or
% paths for which there is nothing more to say than that they exist.

% Timelines: Often objects in a domain must be described by exactly one
% token for every given timepoint in the plan. Such objects are so
% common that we provide a special construct in \eu to extend these
% semantics to derived classes. Any instances of a class derived from a
% Timeline will induce ordering requirements among its tokens in order
% to ensure no temporal overlap may occur among them. See below for
% further discussion.

% Resources: Metric resources, e.g. the energy of a battery or the
% capacity of a cargo hold, are objects with an explicit quantitative
% state in time and with a circumscribed range of changes that can occur
% to impact that state i.e. produce, consume, use, change. These changes
% are captured as tokens. Resources are such a common requirement for
% \eu users that special constructs are also provided for
% them. Instances of classes derived from a Resource will induce
% ordering requirements on their Tokens in order to ensure that the
% level of the resource remains within specified limits.

\item[\textbf{Timelines}] It may be sufficient to maintain only a
  partial-order among tokens. However, it is often the case that
  tokens represent states and actions of a single object in the
  system. Such tokens are typically mutually-exclusive. \eu uses a
  Timeline structure \cite{mus94} to concisely capture system
  components whose behavior is described over time in this manner.

\end{description}

For example, consider the tire-world domain. The tire was located in
the trunk with the predicate tireLocated(Trunk) and located on the
ground with the predicate tireLocated(Ground). Clearly, the same tire
cannot be in both places at once. A Timeline provides a simple method
for aggregating the statements about a tire such that they are
mutually exclusive.

Figure 1: A Timeline for a Tire

Figure 1 illustrates how a Timeline can be used to describe the
whereabouts of a tire. The tire is an object in the tire-world
represented as a timeline. It has predicates associated with it which
can describe its states and actions over time. The predicate names
need not contain the tire prefix; this is now implicit since the
tokens are assigned to a specific tire instance (i.e. an instance of a
Timeline). While Timelines are a useful element of the \eu planning
paradigm, they are not essential. A more general notion of a system
Object can be used which does not impose restrictions of
mutual-exclusion or non-zero duration. This can be important in
supporting partial-order planning.

For this example, the state of the tire is specified using 3
contiguous tokens. The end and start time-points are related by an
equality constraint. A Moving predicate has been introduced to cover
the transition from one location to another. It takes 2 location
arguments. Notice that precedence relationships exist between tokens
such that they cannot overlap but the start and end times may remain
flexible. To illustrate this, sample times are included. Assume a
total time range of interest between 0 and 1000. In addition, tokens
on Timelines have a minimum duration of 1. As a result the values
shown are the most flexible possible values for the start and end of
each token. There are a number of advantages of allowing this
flexibility. First, the basic structure of the plan can be developed
without over-committing to specific times. If it is not necessary for
the tire to be on the ground at time-step 4, then a planner should not
be forced to specify it. Such an approach permits a least-commitment
approach to planning. Second, in many domains it is simply impossible
to know in planning exactly how long an activity might take. For
example, a common activity of driving a car from one point to another
on a road can take varying amounts of time depending on traffic and
traffic lights. In such circumstances it is more practical to express
upper and lower bounds on durations which naturally lead to intervals
for start and end times. In such domains, flexibility in planning aids
robustness in execution.  Rules

In order for a plan to be valid, it must comply with all rules and
regulations pertinent to the application domain in question. Rules
govern the internal and external relationships of a token. For
example, consider a parameterized predicate describing a transition
from one location to another. Let the parameters be from and to. The
parameters are instantiated on a token as variables whose domain of
values is the set of all locations in a given problem. A rule
governing an internal relationship among token variables might
stipulate that a transition must involve a change in location. This
can be easily expressed as a constraint on the definition of a
predicate of the form: from != to. It is reasonable to further
stipulate that one must be Located somewhere before a transition can
occur, and one must end up Located somewhere when completed. This is
an example of a rule governing an external relationship among
tokens. It specifies a requirement that tokens of the predicate
Located precede and succeed tokens of the predicate Moving.

Figure 2: Internal and external relationships from rules on Moving
Figure 2 illustrates the entities and relations involved in specifying
such a rule on a Moving predicate. The token on which the rule applies
is referred to as the master. Each Located token required by the
master is referred to as a slave. All variables are indicated by name
and their domains are expressed as intervals in the case of temporal
variables and as enumerations for the remainder. Application of a rule
on a token can thus cause slave tokens, variables, and constraints to
be introduced.  Token State Model

The capability of a domain rule to cause a slave token to be created
is a key vehicle through which planning occurs. Semantically, this
rule imposes a requirement for supporting tokens to be in the plan in
order for the master to be valid. There are 2 possibilities to
consider:

The slave is inserted as an active token in the plan. As such, rules
may be activated on the slave, and it may consume resources.

The token is merged with a matching token already in the plan. Once a
slave is merged, the requirement it represents are considered
satisfied. The process of merging passes on all restrictions imposed
on the slave to the active token upon which it is merged. Merging a
token requires finding a target active token that is compatible with
the inactive token. For an active token and an inactive token to be
compatible requires that they are instances of the same predicate and
that no intersections between corresponding variables are empty. The
effect of merging is illustrated in Figure 3.

Figure 3: Merging an inactive token on an active token. Domain
restrictions occur in highlighted variables of active token.

On creation of a token, where the commitment has not been made yet to
activate or merge the token, the token is said to be inactive. There
is a nuance to the state model for tokens which relate to the mode of
its creation. If the token is allocated explicitly by an external
actor, rather than internally through rule firing, it may include the
state rejected indicating the planner is permitted to reject the
token. A valid plan can include rejected tokens. This typically arises
where the token represents a goal that is preferable to achieve but
not mandatory. This state is not reachable for slaves since that would
imply selective adherence to the domain model. Figure 4 presents the
state transition diagram for token states relevant for planning. The
transitions are operations on a partial plan which can decide an
outcome for an inactive token. As operations which may arise in
search, they must be reversible during backtracking. The cancellation
operations in each case are also shown.

Each of the token variables, including the parameter variables, has a
domain of values assigned to it.  The variables may also participate
in constraints that specify which value combinations are valid.  To
allow the specification of multiple values, e.g, to express a range of
possible start times, variables are used to specify parameter, start
and end time values for a token.  As a result, a token $T$ is a tuple
$\langle v, P, s, e \rangle$, where $v$ is a variable denoting a state
variable, $P$ is the name of a procedure and $s$ and $e$ (satisfying
$s\leq e$) are variables encoded as flexible token start and end time
points. Consequently, partial plans represent a \textit{range of
  possible solutions} rather than an explicit definition of a single
trajectory of state.  Fig. \ref{fig:flex-timelines} shows an example
of a part of such plans with tokens representing constraints on
flexible temporal intervals. Systematic computation is reflected by
the use of the \texttt{Allen Algebra} \cite{allen84} temporal
relations shown in Fig. \ref{fig:allen-algebra}. It is the use of such
interval arithmetic in the \eu domain model that allows
encoding of temporal constraint relationships between tokens on the
same or other concurrent timelines. Details of propagation algorithms
or of interval logic are beyond the scope of this paper.

@@

\begin{minipage}[c]{\textwidth}
\vspace{+0.5cm}
% \framebox[\textwidth][t]{
Volume\_Survey(x$_1$,y$_1$,x$_n$,y$_n$,Min\_Depth, Max\_Depth)\\
$\Rightarrow$ \\
met\_by Go(x$_1$,y$_1$,Min\_Depth); \\
starts [50,0] Go(x$_2$,y$_2$,Min\_Depth,Max\_Depth); \\
\ldots{} \\
ends [50,0] Go(x$_{n-1}$,y$_{n-1}$, Min\_Depth,Max\_Depth); \\
meets Go(x$_n$,y$_n$,Min\_Depth, Max\_Depth); 
\end{minipage}

%*********

Timelines

It may be sufficient to maintain only a partial-order among
tokens. However, it is often the case that tokens represent states and
actions of a single object in the system. Such tokens are typically
mutually-exclusive. \eu uses a Timeline structure developed in (??)
to concisely capture system components whose behavior is described
over time in this manner.

\comment{This is the first instance that this domain is being
  introduced. Either we need more explanation or it ought to be
  removed.} For example, consider the tire-world domain. The tire was
located in the trunk with the predicate tireLocated(Trunk) and located
on the ground with the predicate tireLocated(Ground). Clearly, the
same tire cannot be in both places at once. A Timeline provides a
simple method for aggregating the statements about a tire such that
they are mutually exclusive.

Figure 1: A Timeline for a Tire

Figure 1 illustrates how a Timeline can be used to describe the
whereabouts of a tire. The tire is an object in the tire-world
represented as a timeline. It has predicates associated with it which
can describe its states and actions over time. The predicate names
need not contain the tire prefix; this is now implicit since the
tokens are assigned to a specific tire instance (i.e. an instance of a
Timeline). While Timelines are a useful element of the \eu planning
paradigm, they are not essential. A more general notion of a system
Object can be used which does not impose restrictions of
mutual-exclusion or non-zero duration. This can be important in
supporting partial-order planning.

For this example, the state of the tire is specified using 3
contiguous tokens. The end and start time-points are related by an
equality constraint. A Moving predicate has been introduced to cover
the transition from one location to another. It takes 2 location
arguments. Notice that precedence relationships exist between tokens
such that they cannot overlap but the start and end times may remain
flexible. To illustrate this, sample times are included. Assume a
total time range of interest between 0 and 1000. In addition, tokens
on Timelines have a minimum duration of 1. As a result the values
shown are the most flexible possible values for the start and end of
each token. There are a number of advantages of allowing this
flexibility. First, the basic structure of the plan can be developed
without over-committing to specific times. If it is not necessary for
the tire to be on the ground at time-step 4, then a planner should not
be forced to specify it. Such an approach permits a least-commitment
approach to planning. Second, in many domains it is simply impossible
to know in planning exactly how long an activity might take. For
example, a common activity of driving a car from one point to another
on a road can take varying amounts of time depending on traffic and
traffic lights. In such circumstances it is more practical to express
upper and lower bounds on durations which naturally lead to intervals
for start and end times. In such domains, flexibility in planning aids
robustness in execution.  Rules

In order for a plan to be valid, it must comply with all rules and
regulations pertinent to the application domain in question. Rules
govern the internal and external relationships of a token. For
example, consider a parameterized predicate describing a transition
from one location to another. Let the parameters be from and to. The
parameters are instantiated on a token as variables whose domain of
values is the set of all locations in a given problem. A rule
governing an internal relationship among token variables might
stipulate that a transition must involve a change in location. This
can be easily expressed as a constraint on the definition of a
predicate of the form: from != to. It is reasonable to further
stipulate that one must be Located somewhere before a transition can
occur, and one must end up Located somewhere when completed. This is
an example of a rule governing an external relationship among
tokens. It specifies a requirement that tokens of the predicate
Located precede and succeed tokens of the predicate Moving.

Figure 2: Internal and external relationships from rules on Moving
Figure 2 illustrates the entities and relations involved in specifying
such a rule on a Moving predicate. The token on which the rule applies
is referred to as the master. Each Located token required by the
master is referred to as a slave. All variables are indicated by name
and their domains are expressed as intervals in the case of temporal
variables and as enumerations for the remainder. Application of a rule
on a token can thus cause slave tokens, variables, and constraints to
be introduced.  Token State Model

The capability of a domain rule to cause a slave token to be created
is a key vehicle through which planning occurs. Semantically, this
rule imposes a requirement for supporting tokens to be in the plan in
order for the master to be valid. There are 2 possibilities to
consider:

The slave is inserted as an active token in the plan. As such, rules
may be activated on the slave, and it may consume resources.

The token is merged with a matching token already in the plan. Once a
slave is merged, the requirement it represents are considered
satisfied. The process of merging passes on all restrictions imposed
on the slave to the active token upon which it is merged. Merging a
token requires finding a target active token that is compatible with
the inactive token. For an active token and an inactive token to be
compatible requires that they are instances of the same predicate and
that no intersections between corresponding variables are empty. The
effect of merging is illustrated in Figure 3.

Figure 3: Merging an inactive token on an active token. Domain
restrictions occur in highlighted variables of active token.

On creation of a token, where the commitment has not been made yet to
activate or merge the token, the token is said to be inactive. There
is a nuance to the state model for tokens which relate to the mode of
its creation. If the token is allocated explicitly by an external
actor, rather than internally through rule firing, it may include the
state rejected indicating the planner is permitted to reject the
token. A valid plan can include rejected tokens. This typically arises
where the token represents a goal that is preferable to achieve but
not mandatory. This state is not reachable for slaves since that would
imply selective adherence to the domain model. Figure 4 presents the
state transition diagram for token states relevant for planning. The
transitions are operations on a partial plan which can decide an
outcome for an inactive token. As operations which may arise in
search, they must be reversible during backtracking. The cancellation
operations in each case are also shown.

Figure 4: Token States and Transitions for Planning

The state of a token is embodied with a 5th and final built-in
variable referred to as the state variable. The planner states are
values in the domain of this variable {MERGED, ACTIVE, REJECTED}. The
INACTIVE state is captured by the variable being unbound. If a value
is removed from the domain, that state will not be reachable. For
example, when a slave is created via a rule, the REJECTED value is
removed from its state variable.

% \textit{Summary}

% This has presented the main elements of the EUROPA planning approach. They are:

% \begin{enumerate}

% \item Temporally scoped predicates and actions (a.k.a. Tokens) to represent
% states and actions in time. Note that tokens do not discriminate
% between state and action.

% \item Token States which are the basis of planning operations on a partial
% plan.

% \item Constraints to describe relationships among tokens. This provides an
% expressive method of describing interactions among tokens in the
% context of Temporal Planning.

% \item Timelines as a concise abstraction to express the evolution of state
% and behavior for a system component. It provides semantics of mutual
% exclusion sequencing in time. Other core abstractions are available in
% \eu for handling metric resources.

% \item Domain rules to describe internal and external relationships on and
% between tokens respectively. Rules are applied to active tokens
% (master tokens) referred to as masters and typically produce inactive
% tokens (slaves) which are a key vehicle through which the planning
% process evolves. The rule structure in \eu differs from the more
% restrictive commitment to preconditions and effects in classical
% planning which prohibits durative actions and disembodied effects
% (i.e. effects which may not occur until some temporal distance after
% the end of the action).

% \end{enumerate}
